A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-09-01 arxiv/transfer.tex
edit

\section{Special Case: Isotropic Covariance and Covariate Shift}\label{sec_special}

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
The results in Section \ref{sec_general} provide conditions for when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
Next, we relate the general conditions to specific properties of task data.
In particular, we consider an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
We show theoretical implications for understanding the performance of multi-task learning.
(a) \textit{Task similarity}: we explain the phenomenon of negative transfer precisely as tasks become more different.
(b) \textit{Sample ratio}: we further explain a curious phenomenon where increasing the sample ratio helps initially, but hurts eventually.
(c) \textit{Covariate shift}: as the source sample size increases, we show that the covariate shift worsens the performance of the multi-task estimator.
Finally, we extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.

@@

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness.
We formalize this connection in the following simplified setting, where we can perform explicit calculations.
We show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	Assume that for  task two, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	For our purpose, it is enough to think of the order of $d$ being $1/\sqrt{p}$ and $pd^2/\sigma^2$ being constant. The more precise conditions on the relations between $d^2$, $\sigma^2$ and $\kappa^2$ are given in \eqref{choiceofpara}.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

@@

>>>>

\section{Interpreting Bias and Variance in an Isotropic Covariance Setting}\label{sec_special}

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
The results in Section \ref{sec_general} provide conditions for when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
Next, we relate the general conditions to specific properties of task data.
In particular, we consider an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
We show theoretical implications for understanding the performance of multi-task learning.
(i) \textit{Task similarity}: we explain the phenomenon of negative transfer precisely as tasks become more different.
(ii) \textit{Sample ratio}: we further explain a curious phenomenon where increasing the sample ratio helps initially, but hurts eventually.
(iii) \textit{Covariate shift}: as the source sample size increases, we show that the covariate shift worsens the performance of the multi-task estimator.
%Finally, we provide a extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.

\smallskip
\noindent\textbf{The isotropic covariance setting.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	Assume that for  task two, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	For our purpose, it is enough to think of the order of $d$ being $1/\sqrt{p}$ and $pd^2/\sigma^2$ being constant.
%The more precise conditions on the relations between $d^2$, $\sigma^2$ and $\kappa^2$ are given in  \eqref{choiceofpara}.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

@@

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness.
We formalize this connection in the following simplified setting, where we can perform explicit calculations.
We show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
In particular, the bias of MTL increases as the distance between $\beta_1$ and $\beta_2$ increases.
Therefore, while the variance of MTL still reduces compared to STL, if the bias increases more than the amount of variance reduced, we will observe negative transfer.

@@

\begin{figure}[!t]
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.pdf}
		\caption{Task similarity}
		\label{fig_model_shift}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.pdf}
		\caption{Sample size}
		\label{fig_size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.pdf}
		\caption{Covariate shift}
		\label{fig_covariate}
	\end{subfigure}
	\caption{%Three takeaways of our theory in Section \ref{sec_insight}.
	We observe a transition from positive to negative transfer as (a) \textit{task model distance} increases and (b) source/target \textit{sample ratio} increases.
	For the special case of having the same task model, we observe in (c) that as source/target \textit{sample ratio} increases, having \textit{covariate shift} worsens the performance of MTL.
	The $y$-axis measures the loss of STL minus MTL.}
	\label{fig_model_shift_phasetrans}
\end{figure}
